---
title: "Integrating JavaScript Libraries into R for Dynamic Visualizations"
author: "your name here"
date: May 23, 2023
execute:
  message: false
  warning: false
format:
  html:
    toc: true
    toc-location: left
    theme: sandstone
---

# 0. Setting up

```{r libraries-and-data}
# import packages ----
library(tidyverse) # data wrangling & viz (with {ggplot2})
library(plotly) # JS plots!
library(DT) # JS tables!
library(leaflet) # JS maps!
library(leaflet.extras) # leaflet add-ons!

# reading in the data ----
lobs <- readRDS(file = here::here("data", "lobsters.rds"))
```

# 1. Summarizing the data

```{r creating-lobs-summary}
# creating new data frame ----
lobs_summary <- lobs %>% 
  
  # calculate total lobster counts by protection status, site, & year (each point will represent lobster counts at a single site for each year from 2012-2018) ----
  group_by(protection_status, site, year) %>% 
  count()
```

# 2. `plotly`

## a. create a static plot

```{r static-boxplot}
static <- lobs_summary %>% 
  
  # create boxplot of mpa vs non-mpa lobster counts ----
  ggplot(aes(x = protection_status, y = n)) +
  
  # geoms: a boxplot and points with jitter ----
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(aes(color = site, shape = site), size = 4, alpha = 0.8, 
             # turn the points into a jitter (with a little more control than geom_jitter)
             position = position_jitter(width = 0.25, height = 0, seed = 1)) +
  
  # update colors and shapes ----
  scale_color_manual(values = c("NAPL" = "#91B38A", 
                                "IVEE" = "#9565CC", 
                                "AQUE" = "#CCC065", 
                                "MOHK" = "#658ACC", 
                                "CARP" = "#CC6565")) +
  scale_shape_manual(values = c(15, 25, 17, 18, 19)) +
  
  # update labels ----
  labs(x = "Protection Status",
       y = "Lobster Counts",
       color = "Site", 
       shape = "Site") + 
  
  # theme ----
  theme_linedraw() +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11))

# print plot ----
static
```

## b. create an interactive plot

```{r basic-interactive-plot}

```

## c. create a better interactive plot

### i. create a marker

```{r summary-marker-df}

```

### ii. make a new static plot with `text = marker` aesthetic argument

```{r new-static-plot, warning = TRUE}
# creating a new static plot ----
static_with_marker <- lobs_summary_marker %>% 
  
  # create boxplot of mpa vs non-mpa lobster counts ----
  ggplot(aes(x = protection_status, y = n, text = marker, group = protection_status)) +
  
  # geoms: boxplot and jitter ----
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(aes(color = site, shape = site), size = 4, alpha = 0.8, 
             position = position_jitter(width = 0.25, height = 0, seed = 1)) +
  
  # update colors and shapes ----
  scale_color_manual(values = c("#91B38A", "#9565CC", "#CCC065", "#658ACC", "#CC6565")) +
  scale_shape_manual(values = c(15, 25, 17, 18, 19)) +
  
  # update labels ----
  labs(x = "Protection Status",
       y = "Lobster Counts",
       color = "Site", 
       shape = "Site") + 
  
  # theme ----
  theme_linedraw() +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11))

# running the ggplot object will give you a scary warning - that's ok! ----
static_with_marker
```

### iii. create plot with markers

```{r better-interactive-plot}

```

### iv. doing things in `plot_ly`

```{r doing-in-plotly}
plot_ly(
  
  # call the data ----
  lobs_summary_marker,
  
  # axes ----
  x = ~ protection_status,
  y = ~ n,
  
  # type: plot_ly equivalent of "geom" ----
  type = "box",
  
  # show underlying data ----
  boxpoints = "all",
  
  # center points on boxplot ----
  pointpos = 0,
  
  # control width of jitter ----
  jitter = 0.25,
  
  # tooltip ----
  hoverinfo = "text", 
  text = ~ marker,
  
  # colors ----
  color = ~ protection_status,
  colors = c("cornflowerblue", "darkgreen")) %>% 
  
  # layout: most formatting goes here! ----
  layout(
    
    # global font option 
    font = list(family = "Times", 
                size = 14),
    
    # changing axis labels 
    xaxis = list(title = list(text = "Protection status")),
    yaxis = list(title = list(text = "Lobster count")),
    
    # editing the marker/tooltip/hoverlabel 
    hoverlabel = list(
      # editing the font: all goes in a list()
      font = list(
        family = "Times",
        size = 13,
        color = "#FFFFFF",
        align = "left"
      )
    )
  )
```

# 3. `DT`

## a. create a basic interactive table

```{r basic-DT}

```

## b. customizing your DT

```{r advanced-DT}

```

# 4. `leaflet`

## a. some cleaning and filtering

```{r unique-sites}
# create df of unique sites ----
sites <- lobs %>% 
  select(site, protection_status, lat, lon) %>% 
  distinct()

# just mpa sites ----
mpa <- sites %>% 
  filter(protection_status == "MPA")

# just non-mpa sites ----
non_mpa <- sites %>% 
  filter(protection_status == "non-MPA")
```

## b. create a custom icon

```{r lobster-icon}

```

## c. create a map!

```{r leaflet-map}

```

# 4. saving the html objects

```{r saving-map, eval = FALSE}
htmltools::save_html(lobs_interactive, here::here("outputs", "plotly.html"))
htmltools::save_html(lobs_dt, here::here("outputs", "dt.html"))
htmltools::save_html(site_map, here::here("outputs", "leaflet.html"))
```

# 5. Prepare a report with your interactive visualizations!

Dynamic visualizations are particularly effective when embedded in reports – check out the fictitious reports (one `.qmd` and `.rmd`, for a side-by-side comparison) in the `example-reports/` folder in this repo:

-  `report-qmd.qmd`
-  `report-rmd.rmd`

Click the **Knit** or **Render** button on the open report file(s) to convert the markdown and code into a clean, organized, and easy-to-read HTML file.

While you may notice some feature and formatting differences between the two, both Quarto documents and R Markdown documents are effective tools for generating reproducible reports that combine prose, code, and outputs.