---
title: "Integrating JavaScript Libraries into R for Dynamic Visualizations"
author: "your name here"
date: "May 23, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 0. Setting up

```{r libraries-and-data, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
library(tidyverse) # data wrangling & viz (with {ggplot2})
library(plotly) # JS plots!
library(DT) # JS tables!
library(leaflet) # JS maps!
library(leaflet.extras) # leaflet add-ons!

# reading in the RDS (saved data object)
lobs <- readRDS(file = here::here("data", "lobsters.rds"))
```

**Note:** RDS (R Data Serialization) is a data file format commonly used for saving R objects. RDS files are relatively small, take less time to import/export, and preserve data types and classes (e.g. factors and dates), eliminating the need to redefine data types after loading the file. We've done some pre-processing of the original/raw data and wrote it out as an `.rds` file (using `saveRDS()`) for us to use in this workshop.

# 1. Summarizing the data

```{r creating-lobs-summary, eval = TRUE, echo = TRUE}
# creating new data frame
lobs_summary <- lobs %>% 
  
  # calculate total lobster counts by protection status, site, & year (each point will represent lobster counts at a single site for each year from 2012-2018) ----
  group_by(protection_status, site, year) %>% 
  count()
```

# 2. `plotly`

## a. create a static plot

```{r static-boxplot, eval = TRUE, echo = TRUE}
static <- lobs_summary %>% 
  
  # create boxplot of mpa vs non-mpa lobster counts ----
  ggplot(aes(x = protection_status, y = n)) +
  
  # geoms: a boxplot and points with jitter
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(aes(color = site, shape = site), size = 4, alpha = 0.8, 
             # turn the points into a jitter (with a little more control than geom_jitter)
             position = position_jitter(width = 0.25, height = 0, seed = 1)) +
  
  # update colors ----
  scale_color_manual(values = c("NAPL" = "#91B38A", 
                                "IVEE" = "#9565CC", 
                                "AQUE" = "#CCC065", 
                                "MOHK" = "#658ACC", 
                                "CARP" = "#CC6565")) +
  scale_shape_manual(values = c(15, 25, 17, 18, 19)) +
  
  # update labels ----
  labs(x = "Protection Status",
       y = "Lobster Counts",
       color = "Site", 
       shape = "Site") + 
  
  # theme ----
  theme_linedraw() +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11))

# print plot ----
static
```

## b. create an interactive plot

```{r basic-interactive-plot, eval = TRUE, echo = TRUE}
ggplotly(static) # ta-da!
```

## c. create a better interactive plot

### i. create a marker

```{r summary-marker-df, eval = TRUE, echo = TRUE}
# adding a column to lobs_summary 
lobs_summary_marker <- lobs_summary %>% 
  
  # create a new column called "marker" ----
  mutate(marker = paste("Site:", site, "<br>",
                      "Year:", year, "<br>",
                      "Status:", protection_status, "<br>",
                      "Lobster count:", n))
```

### ii. make a new static plot with `text = marker` aesthetic argument

```{r new-static-plot, eval = TRUE, echo = TRUE, warning = TRUE}
# creating a new static plot
static_with_marker <- lobs_summary_marker %>% 
  
  # create boxplot of mpa vs non-mpa lobster counts ----
  ggplot(aes(x = protection_status, y = n, text = marker, group = protection_status)) +
  
  # geoms: boxplot and jitter
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(aes(color = site, shape = site), size = 4, alpha = 0.8, 
             position = position_jitter(width = 0.25, height = 0, seed = 1)) +
  
  # update colors ----
  scale_color_manual(values = c("#91B38A", "#9565CC", "#CCC065", "#658ACC", "#CC6565")) +
  scale_shape_manual(values = c(15, 25, 17, 18, 19)) +
  
  # update labels ----
  labs(x = "Protection Status",
       y = "Lobster Counts",
       color = "Site", 
       shape = "Site") + 
  
  # theme ----
  theme_linedraw() +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11))

# printing the ggplot object will give you a scary warning - that's ok! ----
static_with_marker
```

### iii. create plot with markers

```{r better-interactive-plot, eval = TRUE, echo = TRUE}
# tooltip = "text" corresponds to the aes() text call from above!
lobs_interactive <- ggplotly(static_with_marker, tooltip = "text") %>% 
  
  # layout: most formatting goes here! ----
  layout(
    font = list(family = "Times"),
  
    # editing the marker/tooltip/hoverlabel
    hoverlabel = list(
      # editing the font: all goes in a list()
      font = list(
        family = "Times",
        size = 13,
        color = "#FFFFFF",
        align = "left"
      )
    )
  )

# print plot ----
lobs_interactive
```

### iv. doing things in `plot_ly`

```{r doing-in-plotly, eval = TRUE, echo = TRUE}
plot_ly(
  
  # call the data ----
  lobs_summary_marker,
  
  # axes ----
  x = ~ protection_status,
  y = ~ n,
  
  # type: plot_ly equivalent of "geom" ----
  type = "box",
  
  # show underlying data ----
  boxpoints = "all",
  
  # center points on boxplot ----
  pointpos = 0,
  
  # control width of jitter ----
  jitter = 0.25,
  
  # tooltip ----
  hoverinfo = "text", 
  text = ~ marker,
  
  # colors ----
  color = ~ protection_status,
  colors = c("cornflowerblue", "darkgreen")) %>% 
  
  layout(
    # global font option
    font = list(family = "Times", 
                size = 14),
    
    # changing axis labels
    xaxis = list(title = list(text = "Protection status")),
    yaxis = list(title = list(text = "Lobster count")),
    
    # editing the marker/tooltip/hoverlabel
    hoverlabel = list(
      # editing the font: all goes in a list()
      font = list(
        family = "Times",
        size = 13,
        color = "#FFFFFF",
        align = "left"
      )
    )
  )
```

# 3. `DT`

## a. create a basic interactive table

```{r basic-DT, eval = TRUE, echo = TRUE}
datatable(data = lobs)
```

## b. customizing your DT

```{r advanced-DT, eval = TRUE, echo = TRUE}
lobs_dt <- datatable(data = lobs, 
                     # make the column names nice ----
                     colnames = c("Year", "Date", "Site", "Protection status", "Transect", "Replicate", 
                                  "Size (mm)", "Count", "Latitude", "Longitude"),
                     
                     # column filters: sliders and drop down menus ----
                     filter = "top", 
                     
                     # extensions: lots of these! ----
                     extensions = c("Buttons", "ColReorder"),
                     
                     # options ----
                     options = list(
                       # list 10 entries at once
                       pageLength = 10, 
                       # automatically size columns
                       autoWidth = TRUE,
                       # highlight entries that match search term
                       searchHighlight = TRUE,
                       # allow regular expressions and case insensitive searches
                       search = list(regex = TRUE, caseInsensitive = TRUE),
                       dom = "Bfrtip",
                       # buttons options
                       buttons = c("copy", "csv", "excel", "pdf", "print", "colvis"),
                       # links to extension call
                       colReorder = TRUE)
) %>% 
  
  # styling cells: coloring site background ----
formatStyle(
  "site",
  # styleEqual allows matches to column contents ----
  backgroundColor = styleEqual(
    levels = list("NAPL", "IVEE", "AQUE", "MOHK", "CARP"),
    values = c("NAPL" = "#91B38A", 
               "IVEE" = "#9565CC", 
               "AQUE" = "#CCC065", 
               "MOHK" = "#658ACC", 
               "CARP" = "#CC6565")
  )
)

lobs_dt
```

# 4. `leaflet`

## a. some cleaning and filtering

```{r unique-sites, eval = TRUE, echo = TRUE}
# create df of unique sites ----
sites <- lobs %>%  
  select(site, protection_status, lat, lon) %>%  
  distinct()

# just mpa sites ----
mpa <- sites %>%  
  filter(protection_status == "MPA")

# just non-mpa sites
non_mpa <- sites %>%  
  filter(protection_status == "non-MPA")
```

## b. create a custom icon

```{r lobster-icon, eval = TRUE, echo = TRUE}
lobster_icon <- makeIcon(
  iconUrl = "../media/lobster.png",
  iconWidth = 30, iconHeight = 30
)
```
**Note:** Above, we specify the file path to our lobster image, relative to this file (we need to back out a directory level, from `/keys` to `SORTEE-May2023`, then into `media/lobster.png`)

## c. create a map!

```{r leaflet-map, eval = TRUE, echo = TRUE}
# initialize map ----
site_map <- leaflet() %>%  
  
  # add base map tiles (use `addTiles()` for Google maps tiles, OR `addProviderTiles()` for 3rd party base maps: https://leaflet-extras.github.io/leaflet-providers/preview/) ----
  addProviderTiles(providers$Esri.WorldImagery, group = "ESRI World Imagery") %>%  # modify base map appearance with options: `providerTileOptions(opacity = 0.5, noWrap = FALSE)`
  addProviderTiles(providers$Esri.OceanBasemap, group = "ESRI Oceans") %>%  
  
  # add mini map ----
  addMiniMap(toggleDisplay = TRUE, minimized = TRUE) %>%  
  
  # set view over Santa Barbara Channel ----
  setView(lng =  -119.83, lat = 34.44, zoom = 9) %>%  
  
  # add MPA markers ----
  addMarkers(data = mpa, group = "MPA Sites",
             icon = lobster_icon,
             lng = ~lon, lat = ~lat,
             popup = paste("Site Name:", mpa$site, "<br>",
                           "Coordinates (lat/long):", mpa$lat, ",", mpa$lon)) %>% 
  
  # add non-MPA markers ----
  addMarkers(data = non_mpa, group = "Non-MPA Sites",
             icon = lobster_icon,
             lng = ~lon, lat = ~lat,
             popup = paste("Site Name:", non_mpa$site, "<br>",
                           "Coorinates (lat/long):", non_mpa$lat, ",", non_mpa$lon)) %>% 

  # add layers control (toggle base map tiles & markers based on group IDs) ----
  addLayersControl(
    baseGroups = c("ESRI World Imagery", "ESRI Oceans"),
    overlayGroups = c("MPA Sites", "Non-MPA Sites")) %>%  
  
  # add reset map button ----
  leaflet.extras::addResetMapButton()

# print map ----
site_map
```

# 4. saving the html objects

```{r saving-map, eval = FALSE, echo = TRUE}
htmltools::save_html(lobs_interactive, here::here("outputs", "plotly.html"))
htmltools::save_html(lobs_dt, here::here("outputs", "dt.html"))
htmltools::save_html(site_map, here::here("outputs", "leaflet.html"))
```
